Imputation of incomplete multilevel data

### 2024-03-01 manuscript.qmd

# Outline

- motivation and intended audience
- the mice framework/package
- background info on missingness mechanisms etc.
- case study

<!-- # Intended audience -->

@@

Situations in which it is clear what the missingness mechanism is are rare. If you lose one case file during data collection, you can argue that the missingness of that case is random and can therefore be ignored. But how does that relate to other missingness in the data set? Alternatively, for some data collection efforts it is known that the missingness is not ignorable. For example, when collecting blood pressure, there is often a clear reason why such information is (not) collected. The absense of blood pressure measurements may therefore indicate that there is no reason to assume and measure abnormal bloodpressure, increasing the likelihood of MNAR. If mechanisms can be deduced from the context of a study, such mechanisms are extremely valuable in solving for the incomplete data.

@@

Prerequisites: incomplete dataset and known multilevel modeling strategy (i.e. the most general analytic model to be applied to imputed data).

1. Load the data, make sure the variables are correctly formatted (e.g. numeric clustering variable)
2. Explore the missingness
2. Fix any cluster-level missingness determinately (note that `2l.only` methods do not work if there are inconsistencies, fix those first!)
2. Re-evaluate the missingness
3. Explore bivariate relations and associations with missingness indicators (optionally test the associations?)
4. For each incomplete variable:
  - Determine the imputation method
  - Choose imputation model predictors (see recipe)
5. Set methods vector and predictor matrix
6. Impute the incomplete data
6. If error...
6. After successful run:
  - Evaluate convergence
  - Evaluate imputations
9. Analyze and pool
10. For further multilevel model building, AICs could be pooled, how?

>>>>

@@

Situations in which the missingness mechanism is certain are rare. If you lose one case file during data collection, you can argue that the missingness of that case is random and can therefore be ignored. But how does that relate to other missingness in the data set? Alternatively, for some data collection efforts it is known that the missingness is not ignorable. For example, when collecting blood pressure, there is often a clear reason why such information is (not) collected. The absense of blood pressure measurements may therefore indicate that there is no reason to assume and measure abnormal bloodpressure, increasing the likelihood of MNAR. Solving for missing blood pressure would then too much the observations and could easily lead to biased estimation of the effect. If mechanisms can be deduced from the context of a study, such mechanisms are extremely valuable in solving for the incomplete data.

@@

Prerequisites: incomplete dataset and known multilevel modeling strategy (i.e. the most general analytic model to be applied to imputed data), plus an assumed missingness mechanism. This tutorial assumes M(C)AR, for MNAR see Munoz et al.

1. Load the data, make sure the variables are correctly formatted (e.g. numeric clustering variable)
2. Explore the missingness
2. Fix any cluster-level missingness determinately (note that `2l.only` methods do not work if there are inconsistencies, fix those first!)
2. Re-evaluate the missingness
3. Explore bivariate relations and associations with missingness indicators (optionally test the associations?)
4. For each incomplete variable:
  - Determine the imputation method
  - Choose imputation model predictors (see recipe)
5. Set methods vector and predictor matrix
6. Impute the incomplete data
6. If error...
6. After successful run:
  - Evaluate convergence
  - Evaluate imputations
9. Analyze and pool
9. If error... (e.g. rescale variables by converting imputations to long format, mutate, then back to mids)
10. For further multilevel model building, AICs could be pooled, how?

The most general analytic model to be applied to imputed data (Hox et al, 2018).

\begin{align}
\text{popularity}_{ij} = &\gamma_{00}+ \gamma_{10} \text{gender}_{ij} + \gamma_{20} \text{extraversion}_{ij} + \\
&\gamma_{01} \text{experience}_j + \gamma_{21} \text{extraversion}_{ij} \times \text{experience}_j + \\
&u_{2j} \text{extraversion}_{ij} + u_{0j}+ e_{ij} 
\end{align}

<!-- u_{1j} \text{gender}_{ij} +  -->

Setup `R` environment
```{r}
library(mice)     # for imputation
library(ggmice)   # for visualization
library(ggplot2)  # for visualization
library(dplyr)    # for data manipulation
library(lme4)     # for multilevel modeling
```

Load the data
```{r}
load("data/popular_MAR.RData")
```

Inspect the missingness
```{r}
plot_pattern(popular_MAR)
```

Fix cluster-level missingness determinately
```{r}
meth <- make.method(popular_MAR)
meth["experience_j"] <- "2lonly.mean"
meth[meth != "2lonly.mean"] <- ""
pred <- make.predictorMatrix(popular_MAR)
pred[, "cluster_id"] <- -2
imp <- mice(
  popular_MAR, 
  m = 1, 
  method = meth, 
  predictorMatrix = pred,
  maxit = 1)
plot_pattern(complete(imp))
```

Does not work if there are inconsistencies within clusters! But we can do this manually.

```{r}
popular_MAR |>
  group_by(cluster_id) |>
  reframe(teacher_experience = unique(experience_j))
```

This shows us that in the first cluster teacher experience of 4 and 24 is recorded.

```{r}
popular_MAR |>
  filter(cluster_id == 1) |>
  ggplot(aes(experience_j)) + geom_histogram()
```

For most pupils, teacher experience is 24. We can quite safely assume that the teacher expeience of 4 is a data entry error. We can fix this by over-imputing the teacher experience of 4 to 24, and deductively imputing the missing teacher experience from the same cluster as 24.
